### 2024-07-13 base-rates.qmd
20240713 - updates

Another approach is to leverage Bayes' theorem, using Bayesian updating.

@@

::: {#fig-probabilityNomogram}
![](images/probabilityNomogram.png)

@@

For instance, if the starting probability is .5% and the likelihood ratio is 10 (e.g., sensitivity = .90, specificity = .91: $\text{likelihood ratio} = \frac{\text{sensitivity}}{1 - \text{specificity}} = \frac{.9}{1-.91} = 10$), the updated probability is less than 5%, as depicted in @fig-probabilityNomogramLine.
The [`petersenlab`](https://github.com/DevPsyLab/petersenlab) package [@R-petersenlab] contains the `posttestProbability()` function that estimates the posttest probability of an event, given the pretest probability and the likelihood ratio, or given the pretest probability and the sensitivity and specificity of the test.

@@

::: {#fig-probabilityNomogramLine}
![](images/probabilityNomogram.png)

@@

A probability nomogram calculator can be found at the following link: [http://araw.mede.uic.edu/cgi-bin/testcalc.pl](http://araw.mede.uic.edu/cgi-bin/testcalc.pl) (archived at https://perma.cc/X8TF-7YBX).
The [`petersenlab`](https://github.com/DevPsyLab/petersenlab) package [@R-petersenlab] contains the `nomogrammer()` function that creates a nomogram plot, as adapted from https://github.com/achekroud/nomogrammer:

@@

>>>>

Another approach is to leverage Bayes' theorem, using Bayesian updating and the probability nomogram.

@@

::: {#fig-probabilityNomogram}
![](images/probabilityNomogram.png){width=50%}

@@

For instance, if the starting probability is 0.5% and the likelihood ratio is 10 (e.g., sensitivity = .90, specificity = .91: $\text{likelihood ratio} = \frac{\text{sensitivity}}{1 - \text{specificity}} = \frac{.9}{1-.91} = 10$) from a positive test, the updated probability is less than 5%, as depicted in @fig-probabilityNomogramLine.
The [`petersenlab`](https://github.com/DevPsyLab/petersenlab) package [@R-petersenlab] contains the `posttestProbability()` function that estimates the posttest probability of an event, given the pretest probability and the likelihood ratio, or given the pretest probability and the sensitivity and specificity of the test.

@@

If the starting probability is 0.5% and the likelihood ratio is 0.11 from a negative test, the updated probability is essentially nearly indistinguishable from zero (0.05%).
```{r}
petersenlab::posttestProbability(
  pretestProb = .005,
  likelihoodRatio = 0.11)
```

::: {#fig-probabilityNomogramLine}
![](images/probabilityNomogramLine.png){width=50%}

@@

A probability nomogram calculator can be found at the following link: [http://araw.mede.uic.edu/cgi-bin/testcalc.pl](http://araw.mede.uic.edu/cgi-bin/testcalc.pl) (archived at <https://perma.cc/X8TF-7YBX>).
The [`petersenlab`](https://github.com/DevPsyLab/petersenlab) package [@R-petersenlab] contains the `nomogrammer()` function that creates a nomogram plot, as adapted from Adam Chekroud (<https://github.com/achekroud/nomogrammer>):

@@

The blue line indicates the [posterior probability](#posttestProbability) of the condition given a positive test.
The pink line indicates the [posterior probability](#posttestProbability) of the condition given a negative test.
